Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-09-08 summary_minh_edited.tex

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that a PABM is a special case of a GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. 
This insight allows us to construct new algorithms for community detection
and parameter estimation for the PABM, 
as well as improve an existing algorithm 
that relies on Sparse Subspace Clustering. 
Using established asymptotic properties of 
Adjacency Spectral Embedding for the GRDPG, 
we derive asymptotic properties of these algorithms. 
In particular, we demonstrate that the absolute number of 
community detection errors tends to zero as 
the number of graph vertices tends to infinity. 
Simulation experiments illustrate these properties. 
\end{abstract}

@@

\begin{remark}
  \label{rem:pabm_view2}
% Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a
% graph $G = (V, E)$ with adjacency matrix $A$ 
% such that $A \sim \mathrm{Bernoulli}(P)$.
% Let each vertex have a community label between $1$ and $K$.
% Then $G$ is drawn from a Popularity Adjusted Block Model if
  In a PABM each vertex $i$ has $K$ popularity parameters $\lambda_{i1}, \dots, \lambda_{iK}$, that describe its affinity
toward each of the $K$ communities. % , i.e., vertex $i$ has popularity parameters
% $\lambda_{i1}, ..., \lambda_{iK}$, and  $P_{ij} = \lambda_{i z_j}
% \lambda_{j z_i}$ for all $i \leq j$. 
Another view of a PABM is as follows.
Let $\tilde{P}$ be the matrix obtained by permuting the rows and
columns of $P$ so that the vertices are arranged according to
increasing values of the community assignments $z_i \in
\{1,2,\dots,K\}$. Denote by $\tilde{P}^{(k \ell)}$ the $n_k \times n_{\ell}$
submatrix of $\tilde{P}$ corresponding to the
edge probabilities between vertices in communities
$k$ and $\ell$. Here $n_k = |\{ i \colon z_i = k\}|$ is the number of
vertices assigned to community $k$, for $k = 1,2\dots,K$.
Note that $\tilde{P}^{(k \ell)} = (\tilde{P}^{(\ell k)})^\top$. Next
let $\lambda^{(k \ell)} = \{\lambda_{i \ell} \colon z_i = k\} \in
\mathbb{R}^{n_k}$; the elements of $\lambda^{(k \ell)}$ are the affinity parameters toward the $\ell$th community of
all vertices in the $k$th community. Define $\lambda^{(\ell k)}$
analogously. Then each block $\tilde{P}^{(k \ell)}$ can be written as the outer product of two vectors:
\begin{equation} \label{eq:pabm}
  \tilde{P}^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}.
\end{equation}
% for a set of $K^2$ popularity vectors 
% $\{\lambda^{(st)}\}_{1 \leq s \leq K, 1 \leq t \leq K}$ where each
% $\lambda^{(st)}$ is a column vector of dimension $n_s$.
% . Then a graph $G$
% is drawn from a PABM with parameters
% $\{\lambda^{(st)}\}$ if its adjacency matrix $A$
% satisfies $A \sim \mathrm{BernoulliGraph}(P)$.
We will henceforth use the notation \(A \sim \PABM(\{\lambda^{(k
  \ell)}\}_K, \rho_n)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with $K$ communities, popularity parameters
\(\{\lambda^{(k \ell)}\}\) and sparsity parameter $\rho_n$.
\end{remark}

@@

\begin{definition}[Subspace Detection Property]
  \label{def:subspace_detection}
Let $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$ be noisy
points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the
$y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise
vectors. Let $\lambda \geq 0$ be given and
let $C$ and $B$ be constructed from the
solutions of LASSO problems as described in Eq.~\eqref{eq:ssc} with
this given choice of $\lambda$. Then $X$
is said to satisfy the subspace detection property 
with sparsity parameter $\vartheta$ if each column of
$C$ has nonzero $\ell_2$ norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are
from different subspaces.
\end{definition}

\begin{remark}
  One of the most common approach to show that SSC works for a noisy
  sample $X$ is to show that 
  $X$ satisfies the
  subspace detection property for some choice of $\vartheta$; recall
  that $\vartheta$ is the sparsity parameter $\vartheta$ for
the LASSO problems in Eq.~\eqref{eq:ssc}. However, this is not sufficient to
guarantee that SSC perfectly recovers the underlying subspaces.
More specifically
if $X$ satisfies the
subspace detection property then $B$ describes a graph with {\em at least}
$K$ disconnected subgraphs, with the ideal case being that there are exactly $K$ subgraphs which map onto each
subspace. Nevertheless it is also possible that the $K$ subspaces are represented by
$K' > K$
multiple disconnected subgraphs and we cannot, at least without a subsequent
post-processing step, recover the $K$ subspaces directly from $B$;
see \citet{sdp_sufficiency} and \citet{liu_ssc}
for further discussions. Therefore in practice $B$ is usually treated as an
affinity matrix and, as we allude to earlier, the rows of $B$ are partitioned using some
clustering algorithm to obtain the final clustering. % Finally we note that the subspace detection property is contingent
% on choosing a sufficiently large sparsity parameter $\vartheta$ for
% the LASSO problems in Eq.~\eqref{eq:ssc}.
\end{remark}

@@

We previously stated in Theorem~\ref{theorem2} one possible set of latent positions that result in
the edge probability matrix of a PABM, namely
\(P = \tilde{\Pi} (XU) I_{p, q}
(XU)^\top \tilde{\Pi}^{\top}\) where $X$ is block diagonal and
$\tilde{\Pi}$ is a permutation matrix.  
Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces, i.e. $\langle U^{\top} x_i, U^{\top} x_j \rangle = 0$ whenever $z_i \not = z_j$. 
Thus if
we have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial because \(U\)
is orthonormal and fixed for each value of \(K\). 
However, direct
identification or estimation of \(XU\) is possibly difficult 
due to the non-identifiability of $XU$ (see Remark~\ref{rem:non_identifiable}).
when we are given only $P$. % of the latent
% positions of a GRDPG.  
% due to the
% non-identifiability of \(XU\) when we are only given $P$, 
% direct
% identification or estimation of \(XU\) is possibly difficult 
% if we are only given $P$. 
% (\cite{rubindelanchy2017statistical}).
More specifically, suppose we find a matrix $Y \in
\mathbb{R}^{n \times K^2}$
such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
we  \( Y = \tilde{\Pi} XU Q \) for some indefinite orthogonal matrix
$Q \in \mathbb{O}(p,q)$. %  as there is no
% general matrix decomposition that yield 
% The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
% such that points within each community lie on \(K\)-dimensional
% orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
% removes the orthogonality property but retains the property that each
% community is represented by a \(K\)-dimensional subspace. 
% However, because $\tilde{\Pi}$ is unknown (indeed, if we know $\tilde{\Pi}$
% then we also know the community labels), 
% Therefore, even if we
% start with the true but unknown edge probability matrix $P$ we
% will not be able to recover the latent positions \(XU\).
However since \(Q\) is not necessarily an
orthogonal matrix and hence, if $y_i$ denote the $i$th row of $Y$ then $\langle U^{\top} x_i, U^{\top} x_j \rangle
\neq \langle y_i, y_j \rangle$.
This prevents us from transferring the orthogonality property of
\(XU\) directly to $Y$.

@@

To complete the proof of Theorem~\ref{theorem3}, recall that \(X\) 
is block diagonal with each block corresponding to one community, 
and hence \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community. 
As $B = n VV^{\top} = n \tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top}$, 
we conclude that $B_{ij} = 0$ 
whenever vertices $i$ and $j$ belong to different communities.  
\end{proof}
Under the condition that $\|\Lambda\|_F > 0$, where 
$\Lambda$ is the $n \times K$ matrix of popularity parameters, 
Theorem \ref{theorem3} provides perfect community detection from
\(P\). More specifically let \(|B|\) be the affinity matrix for graph \(G'\), 
where $|\cdot|$ is applied entry-wise. Then
\(G'\) consists of exactly \(K\) disjoint subgraphs, 
as $G'$ has no edges between communities. 
All that is left to identify the communities is 
to assign each subgraph a distinct community label. 
In practice, we do not observe $P$ and instead only observe the noisy
$A \sim \BG(P)$. A natural approach is then to use
the affinity matrix $\hat{B} = n \hat{V} \hat{V}^{\top}$ where
$\hat{V}$ is the matrix of eigenvectors (corresponding to the largest
eigenvalues in modulus) of $A$. The resulting procedure, named
Orthogonal Spectral Clustering, is presented in
Algorithm~\ref{alg:osc}.
The following result leverages existing theoretical properties
of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
$\hat{B}$ converges almost surely to $B$; in particular 
 $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ 
for pair $(i, j)$ in different communities. 
\begin{algorithm}[tp]
  \label{alg:osc}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{Orthogonal Spectral Clustering.}
\end{algorithm}

@@

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
obtained from OSC as described in Algorithm~\ref{alg:osc}. Then for $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4a}
\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. In particular $\hat{B}_{ij} -
B_{ij} \overset{\mathrm{a.s}}{\rightarrow} 0$ where the convergence is
uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
communities we have
% \begin{equation} \label{eq:thm4}
% \max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
% O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
% \end{equation}
% with high probability. 
% Thus for all pairs $(i, j)$ in different communities,
% we have uniform convergence 
$\hat{B}_{ij} \overset{\mathrm{a.s}}{\rightarrow} 0$. 
\end{theorem}
Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of mis-clustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em
  proportion} of mis-clustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

\begin{theorem}
\label{theorem5}
Let $P$ describe the edge probability matrix of the PABM with
$n$ vertices, and let $A \sim \Bernoulli(P)$.  Let $\hat{V}$ be the
matrix of eigenvectors of $A$ corresponding to the $K^2$ largest
eigenvalues in modulus. Then for any $\epsilon > 0$ 
there exists a choice of $\lambda > 0$ and $N \in \mathbb{N}$ such
that for all $n \geq N$, $\sqrt{n} \hat{V}$ obeys the subspace detection property with
probability at least $1 - \epsilon$.  
\end{theorem}

@@

We also note that unlike the MLE-based method
\citep{307cbeb9b1be48299388437423d94bf1}, the ASE method in
Algorithm~\ref{alg:param_est} can be trivially modified so as to not
require the community labels if we are only interested in 
estimating $P$. More specifically we first compute the ASE
$\hat{Z}$ of $A$ (see Remark~\ref{rem:ase}) and then compute $\hat{P}
= \hat{Z} I_{p,q} \hat{Z}^{\top}$. The resulting estimate $\hat{P}$
will have the same convergence rate as that given in Eq.(xxx).
% we obtain by
% combining the popularity vectors outputted by
% Algorithm~\ref{alg:param_est} is asymptotically
% equivalent to using the full ASE of $A$ to reconstruct the edge
% probability matrix.

>>>>

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that the PABM is a special case of 
the GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. 
This insight allows us to construct new algorithms for community detection
and parameter estimation for the PABM, 
as well as improve an existing algorithm 
that relies on Sparse Subspace Clustering. 
Using established asymptotic properties of 
Adjacency Spectral Embedding for the GRDPG, 
we derive asymptotic properties of these algorithms. 
In particular, we demonstrate that the absolute number of 
community detection errors tends to zero as 
the number of graph vertices tends to infinity. 
Simulation experiments illustrate these properties. 
\end{abstract}

@@

\begin{remark}
  \label{rem:pabm_view2}
% Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a
% graph $G = (V, E)$ with adjacency matrix $A$ 
% such that $A \sim \mathrm{Bernoulli}(P)$.
% Let each vertex have a community label between $1$ and $K$.
% Then $G$ is drawn from a Popularity Adjusted Block Model if
In a PABM, each vertex $i$ has $K$ popularity parameters 
$\lambda_{i1}, \dots, \lambda_{iK}$, that describe its affinity
toward each of the $K$ communities. % , i.e., vertex $i$ has popularity parameters
% $\lambda_{i1}, ..., \lambda_{iK}$, and  $P_{ij} = \lambda_{i z_j}
% \lambda_{j z_i}$ for all $i \leq j$. 
Another view of a PABM is as follows.
Let $\tilde{P}$ be the matrix obtained by permuting the rows and
columns of $P$ so that the vertices are arranged according to
increasing values of the community assignments $z_i \in
\{1,2,\dots,K\}$. Denote by $\tilde{P}^{(k \ell)}$ the $n_k \times n_{\ell}$
submatrix of $\tilde{P}$ corresponding to the
edge probabilities between vertices in communities
$k$ and $\ell$. Here $n_k = |\{ i \colon z_i = k\}|$ is the number of
vertices assigned to community $k$, for $k = 1,2\dots,K$.
Note that $\tilde{P}^{(k \ell)} = (\tilde{P}^{(\ell k)})^\top$. Next
let $\lambda^{(k \ell)} = \{\lambda_{i \ell} \colon z_i = k\} \in
\mathbb{R}^{n_k}$; the elements of $\lambda^{(k \ell)}$ are the affinity parameters toward the $\ell$th community of
all vertices in the $k^{th}$ community. Define $\lambda^{(\ell k)}$
analogously. Then each block $\tilde{P}^{(k \ell)}$ can be written as the outer product of two vectors:
\begin{equation} \label{eq:pabm}
  \tilde{P}^{(k \ell)} = \rho_n \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}.
\end{equation}
% for a set of $K^2$ popularity vectors 
% $\{\lambda^{(st)}\}_{1 \leq s \leq K, 1 \leq t \leq K}$ where each
% $\lambda^{(st)}$ is a column vector of dimension $n_s$.
% . Then a graph $G$
% is drawn from a PABM with parameters
% $\{\lambda^{(st)}\}$ if its adjacency matrix $A$
% satisfies $A \sim \mathrm{BernoulliGraph}(P)$.
We will henceforth use the notation \(A \sim \PABM(\{\lambda^{(k
  \ell)}\}_K, \rho_n)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with $K$ communities, popularity parameters
\(\{\lambda^{(k \ell)}\}\) and sparsity parameter $\rho_n$.
\end{remark}

@@

\begin{definition}[Subspace Detection Property]
  \label{def:subspace_detection}
Let $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$ be noisy
points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the
$y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise
vectors. Let $\vartheta \geq 0$ be given and
let $C$ and $B$ be constructed from the
solutions of LASSO problems as described in Eq.~\eqref{eq:ssc} with
this given choice of $\vartheta$. Then $X$
is said to satisfy the subspace detection property 
with sparsity parameter $\vartheta$ if each column of
$C$ has nonzero $\ell_2$ norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are
from different subspaces.
\end{definition}

\begin{remark}
One of the most common approach to show that SSC works for a noisy
sample $X$ is to show that 
$X$ satisfies the
subspace detection property for some choice of $\vartheta$; recall
that $\vartheta$ is the sparsity parameter $\vartheta$ for
the LASSO problems in Eq.~\eqref{eq:ssc}. However, this is not sufficient to
guarantee that SSC perfectly recovers the underlying subspaces.
More specifically, if $X$ satisfies the 
subspace detection property, 
then $B$ describes a graph with {\em at least} $K$ disconnected subgraphs, 
with the ideal case being that there are exactly $K$ subgraphs 
which map onto each subspace. 
Nevertheless it is also possible that the $K$ subspaces are represented by
$K' > K$
multiple disconnected subgraphs and we cannot, at least without a subsequent
post-processing step, recover the $K$ subspaces directly from $B$;
see \citet{sdp_sufficiency} and \citet{liu_ssc}
for further discussions. Therefore in practice $B$ is usually treated as an
affinity matrix and, as we allude to earlier, 
the rows of $B$ are partitioned using some
clustering algorithm to obtain the final clustering. 
\end{remark}

@@

We previously stated in Theorem~\ref{theorem2} one possible set of latent positions that result in
the edge probability matrix of a PABM, namely
\(P = \tilde{\Pi} (XU) I_{p, q}
(XU)^\top \tilde{\Pi}^{\top}\) where $X$ is block diagonal and
$\tilde{\Pi}$ is a permutation matrix.  
Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces, i.e. $\langle U^{\top} x_i, U^{\top} x_j \rangle = 0$ whenever $z_i \not = z_j$. 
Thus if
we have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial because \(U\)
is orthonormal and fixed for each value of \(K\). 
However, direct
identification or estimation of \(XU\) is possibly difficult 
due to the non-identifiability of $XU$ (see Remark~\ref{rem:non_identifiable}).
when we are given only $P$. % of the latent
% positions of a GRDPG.  
% due to the
% non-identifiability of \(XU\) when we are only given $P$, 
% direct
% identification or estimation of \(XU\) is possibly difficult 
% if we are only given $P$. 
% (\cite{rubindelanchy2017statistical}).
More specifically, suppose we find a matrix $Y \in
\mathbb{R}^{n \times K^2}$
such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
we  \( Y = \tilde{\Pi} XU Q \) for some indefinite orthogonal matrix
$Q \in \mathbb{O}(p,q)$. %  as there is no
% general matrix decomposition that yield 
% The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
% such that points within each community lie on \(K\)-dimensional
% orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
% removes the orthogonality property but retains the property that each
% community is represented by a \(K\)-dimensional subspace. 
% However, because $\tilde{\Pi}$ is unknown (indeed, if we know $\tilde{\Pi}$
% then we also know the community labels), 
% Therefore, even if we
% start with the true but unknown edge probability matrix $P$ we
% will not be able to recover the latent positions \(XU\).
However since \(Q\) is not necessarily an
orthogonal matrix and hence, if $y_i$ denote the $i^{th}$ row of $Y$, 
then $\langle U^{\top} x_i, U^{\top} x_j \rangle
\neq \langle y_i, y_j \rangle$.
This prevents us from transferring the orthogonality property of
\(XU\) directly to $Y$.

@@

To complete the proof of Theorem~\ref{theorem3}, recall that \(X\) 
is block diagonal with each block corresponding to one community, 
and hence \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community. 
As $B = n VV^{\top} = n \tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top}$, 
we conclude that $B_{ij} = 0$ 
whenever vertices $i$ and $j$ belong to different communities.  
\end{proof}

Under the condition that $\|\Lambda\|_F > 0$, 
Theorem \ref{theorem3} provides perfect community detection from \(P\).
\footnote{This is excluding the special case in which vectors in each 
subspace lie on mutually orthogonal sub-subspaces.}
More specifically let \(|B|\) be the affinity matrix for graph \(G'\), 
where $|\cdot|$ is applied entry-wise. Then
\(G'\) consists of exactly \(K\) disjoint subgraphs, 
as $G'$ has no edges between communities. 
All that is left to identify the communities is 
to assign each subgraph a distinct community label. 
In practice, we do not observe $P$ and instead only observe the noisy
$A \sim \BG(P)$. A natural approach is then to use
the affinity matrix $\hat{B} = n \hat{V} \hat{V}^{\top}$ where
$\hat{V}$ is the matrix of eigenvectors (corresponding to the largest
eigenvalues in modulus) of $A$. The resulting procedure, named
Orthogonal Spectral Clustering, is presented in
Algorithm~\ref{alg:osc}.
The following result leverages existing theoretical properties
of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
$\hat{B}$ converges almost surely to $B$; in particular 
 $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ 
for each pair $(i, j)$ in different communities.

\begin{algorithm}[tp]
  \label{alg:osc}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{Orthogonal Spectral Clustering.}
\end{algorithm}

@@

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
obtained from OSC as described in Algorithm~\ref{alg:osc}. Then for $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4a}
\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. In particular $\hat{B}_{ij} -
B_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$ where the convergence is
uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
communities we have
% \begin{equation} \label{eq:thm4}
% \max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
% O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
% \end{equation}
% with high probability. 
% Thus for all pairs $(i, j)$ in different communities,
% we have uniform convergence 
$\hat{B}_{ij} \overset{\mathrm{a.s}}{\rightarrow} 0$. 
\end{theorem}
Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of mis-clustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em
  proportion} of mis-clustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

\begin{theorem}
\label{theorem5}
Let $P$ describe the edge probability matrix of the PABM with
$n$ vertices, and let $A \sim \Bernoulli(P)$.  Let $\hat{V}$ be the
matrix of eigenvectors of $A$ corresponding to the $K^2$ largest
eigenvalues in modulus. Then for any $\epsilon > 0$ 
there exists a choice of $\vartheta > 0$ and $N \in \mathbb{N}$ such
that for all $n \geq N$, $\sqrt{n} \hat{V}$ obeys the subspace detection property with
probability at least $1 - \epsilon$.  
\end{theorem}

@@

We also note that unlike the MLE-based method
\citep{307cbeb9b1be48299388437423d94bf1}, the ASE method in
Algorithm~\ref{alg:param_est} can be trivially modified so as to not
require the community labels if we are only interested in 
estimating $P$. More specifically we first compute the ASE
$\hat{Z}$ of $A$ (see Remark~\ref{rem:ase}) and then compute $\hat{P}
= \hat{Z} I_{p,q} \hat{Z}^{\top}$. The resulting estimate $\hat{P}$
will have the same convergence rate as that given in Eq.~(\ref{eq:conv_prob}).
% we obtain by
% combining the popularity vectors outputted by
% Algorithm~\ref{alg:param_est} is asymptotically
% equivalent to using the full ASE of $A$ to reconstruct the edge
% probability matrix.
